---
title: "my_project"
author: "Marcin Solnicki"
date: '2022-12-15'
output: html_document
---

## Short Description

London bike sharing project- the purpose to predict future bike shares.

The project consist of two elements: -exploratory data analysis -advanced analyst

Dataset downloaded from Kaggle [London bike sharing dataset \| Kaggle](https://www.kaggle.com/datasets/hmavrodiev/london-bike-sharing-dataset?datasetId=376751&language=R)

```{r}
library(lattice)
library(dplyr)
library(ggplot2)
df<-read.csv("C:/Users/soma2003/Desktop/bike_share.csv", header = TRUE, sep = ",")
head(df)
```

remove unnecessary variable:

```{r}
df <- subset(df, select = -weather_code)
head(df)
```

Variables:

-   "timestamp" - timestamp field for grouping the data "cnt" - the count of a new bike shares

-   "t1" - real temperature in C

-   "t2" - temperature in C "feels like" "wind_speed"-speed of wind

-   "hum" - humidity in percentage "windspeed" - wind speed in km/h

-   "isholiday" - boolean field - 1 holiday / 0 non holiday

-   "isweekend" - boolean field - 1 if the day is weekend

-   "season" - category field meteorological seasons: 0-spring ; 1-summer; 2-fall; 3-winter.

#extracting year and month from date

```{r}
df$year<-format(as.Date(df$timestamp, format="%Y-%m-%d"), "%Y")
 df$month<-format(as.Date(df$timestamp, format="%Y-%m-%d"), "%m")
tail(df)
```

time frame 2015/01/04- 2017/01/03

```{r}
ggplot(df,aes(x=year,y=cnt))+geom_point()
```

Filtering out 2017 from dataset

```{r}
df<-df%>% filter(year!=2017)
table(df$year)
```

## Histograms

```{r}
par(mfrow=c(2,2))
options(repr.plot.width=12, repr.plot.height=7)
hist(df$cnt, main="Rentals",xlab="rentals");hist
hist(df$t1, main = "Temperature", xlab = "Temperature");
hist(df$hum, main = "Humidity",xlab = "Humidity");
hist(df$wind_speed, main = "Wind speed", xlab = "Wind speed");
par(mfrow=c(1,1))
```

Average Rentals by month

```{r}

avg_cnt_m <- df %>% group_by(month) %>% summarise(avg_cnt = mean(cnt))
ggplot(avg_cnt_m)+geom_col(aes(x=month,y=avg_cnt))+ylab('average bikes ')
```

Summary of whole data

```{r}
summary(df)
```

Rentals By Season:

0-spring,1-summer,2-autumn,3-winter

```{r}
s = c("olivedrab3", 'yellow', 'orange', 'grey50')
options(repr.plot.width=10, repr.plot.height=8)
df %>% group_by(season) %>%
  summarise(n = n(), rent = sum(cnt)) %>%
  ggplot(aes(season, rent, fill = season)) + 
  geom_bar(stat = "identity", show.legend = F, color = 'gray50') +
  labs(title = "Rentals by Season", x = "", y = "Total Rentals") +
 scale_y_continuous(labels = scales::label_comma())
```

```{r}
barchart(~df$cnt|df$is_holiday,main="Rental by holiday or not", xlab="rents")

```

```{r}
barchart(~df$cnt|df$is_weekend,main="Rental by weekend or not",xlab="rentals")
```

Density plot of real temperature grouped by season

```{r}
densityplot(~t1, xlab="temprature", data=df, groups=season, 
plot.points=FALSE, auto.key = TRUE)

```

## Correlation matrix

```{r}
num_d<-select_if(df,is.numeric)
corr<-round(cor(num_d),2)
corr
```

Rentals to humidity(cor=-0.46)

```{r}
ggplot(df,aes(hum,cnt))+geom_point()+ylab("Rentals")+xlab("Humidity")
```

The higher humidity, the less amount of rented bikes. Correlation minus

Cleaning outliers

```{r}
par(mfrow=c(2,2))
boxplot(df$cnt,horizontal = TRUE)
boxplot(df$hum,horizontal = TRUE)
outlh<-boxplot.stats(df$hum)$out
outlc<-boxplot.stats(df$cnt)$out
#outl
df<-df%>%filter(!between(hum,min(outlh),max(outlh)))
df<-df%>%filter(!between(cnt,min(outlc),max(outlc)))
##without outliers
boxplot(df$cnt,horizontal = TRUE)
boxplot(df$hum,horizontal = TRUE)
```

## linear regression

Model summary

```{r}
lm1<-lm(formula = cnt~hum,data=df)
summary(lm1)
```

Model diagnostic

```{r}
options(repr.plot.width = 14,repr.plot.height = 10)
par(mfrow= c(2,2))
plot(lm1)

```

Theoretical (model) vs real values

```{r}
par(mfrow=c(1,1))
plot(df$cnt~df$hum, xlab="hum", ylab="cnt")
abline(lm1, col='red')
legend("topleft", legend=c("real","model"), col=c("black","red"), pch=c(1,NA), lty=c(NA,2), bty="n")
```

Residuals analysis

```{r}

reszty = residuals(lm1)
hist(reszty, prob=TRUE)
lines(density(reszty), col="red", lwd=2)

```

#Insights, Conslusion

-   No missig data or null values

-   July and September are the months with highest average bikes rented

-   Summer is the most popular season

-   Most rented bikes in one-hour 7860

-   The windspeed and season has a very low co-relation with the cnt which means, the windspeed and seasons doesn't have an effect in the demand of bikes rented.

-   Humidity is the most corelated factor with rentals. The higher humidity, the less amount of rented bikes. Correlation minus

-   People rent more bikes during working day than non working day.

-   People don't prefer to ride bikes during holidays.
